---
title: "Human Microbiome Project"
author: "Erin Dahl"
date: '`r format(Sys.Date(), "%B %e, %Y")`'
output: rmarkdown::html_document
vignette: >
  %\VignetteIndexEntry{microshades-HMP}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Human Microbiome Project 16S Data Vignette

This vignette uses 16S rRNA sequencing data from variable region v1-v3 in the Human Microbiome Project.
Stool samples from male subjects will be explored at the Phylum & Family taxonomic levels.


This data is available in the HMP16SData library. To download use `BiocManager::install("HMP16SData")`.

Learn more about the HMP16S Data [here](https://bioconductor.org/packages/release/data/experiment/html/HMP16SData.html)


Additionally, the package [speedyseq](https://github.com/mikemc/speedyseq) is necessary to use the function `prep_mdf()`. The package speedyseq provides faster versions of phyloseq’s plotting and taxonomic merging functions. Alternatively, the phyloseq object can be melted and transformed by using phyloseq functions `tax_glom()` and/or `transform_sample_counts()`, and melted by using `psmelt()`.


## Load the necessary packages for this tutorial

```{r message=FALSE, warning=FALSE}
library(HMP16SData)
library(phyloseq)
library(dplyr)
library(ggplot2)
library(microshades)
library(forcats)
```

## Load and subset the HMP data
```{r}

v1v3_stool <-
    V13() %>%
    subset(select = HMP_BODY_SUBSITE == "Stool") %>%
    subset(select = SEX == "Male")

ps_v1v3_stool <-
    as_phyloseq(v1v3_stool)
```

## Apply microshades functions

Begin using microshades functions to evaluate abundance and apply advanced color organization at the Phylum Family level. 
```{r}
# Use microshades function prep_mdf to agglomerate, normalize, and melt the phyloseq object
mdf_prep <- prep_mdf(ps_v1v3_stool, subgroup_level = "FAMILY")

# Create a color object for the specified data
color_obj_v1v3_stool <- create_color_dfs(mdf_prep, group_level = "PHYLUM", subgroup_level = "FAMILY", cvd = TRUE)


# Extract
mdf_v1v3_stool <- color_obj_v1v3_stool$mdf
cdf_v1v3_stool <- color_obj_v1v3_stool$cdf
```

## Plot 

The dataframe `mdf_v1v3_stool` contains sample data and abundance info. The dataframe `cdf_v1v3_stool` stores the color mapping information used for plotting. 

`plot_microshades` produces a stacked barplot with ordered subgroup taxonomy. The darkest shade is the most abundant, and the lightest shade is the least abundant. (excluding the "other" subgroup from each of the selected groups)

```{r fig.width= 8, fig.height= 4}
plot_1 <- plot_microshades(mdf_v1v3_stool, cdf_v1v3_stool, group_label = "Phylum Family")

plot_1 + scale_y_continuous(labels = scales::percent, expand = expansion(0)) +
  theme(legend.key.size = unit(0.2, "cm"), text=element_text(size=10)) +
  theme(axis.text.x = element_text(size= 6)) 

```

## Reorder Samples/Group levels

The plot above indicates that Bacteroidaceae is the most abundant family in the Bacteriodetes phylum in this dataset. To see samples arranged in order of Bacteroidaceae abundance (or any genus listed in the legend above)  use the `reorder_samples_by()` function and then plot. This function can also reorder the Group levels based on overall group level abundance, if desired steeing the sink_abundant_groups to true.

Reodering the samples can help users notice additional visual trends.

### Reorder the samples by Bacteroidaceae Abundance

```{r fig.width= 8, fig.height= 4}
# reorder_samples_by will change the order of samples based on an abundance of a specified subgroup taxonomy
new_sample_order_v1v3_stool <- reorder_samples_by(mdf_v1v3_stool, cdf_v1v3_stool, order = "Bacteroidaceae", group_level = "PHYLUM", subgroup_level = "FAMILY", sink_abundant_groups = FALSE)

mdf_new_smample_order <-new_sample_order_v1v3_stool$mdf
cdf_new_smample_order <-new_sample_order_v1v3_stool$cdf

plot_2 <- plot_microshades(mdf_new_smample_order, cdf_new_smample_order, group_label = "Phylum Family")

plot_2 + scale_y_continuous(labels = scales::percent, expand = expansion(0)) +
  theme(legend.key.size = unit(0.2, "cm"), text=element_text(size=10)) +
  theme(axis.text.x = element_text(size= 6))

```


### Reorder by Abundance of Group levels Only


```{r fig.width= 8, fig.height= 4}
# reorder_samples_by will change the order of samples based on an abundance of a specified subgroup taxonomy
# The default subgroup_level is "Genus" 
new_group_order_v1v3_stool <- reorder_samples_by(mdf_v1v3_stool, cdf_v1v3_stool, group_level = "PHYLUM", subgroup_level = "FAMILY", sink_abundant_groups = TRUE)

mdf_new_group_order <-new_group_order_v1v3_stool$mdf
cdf_new_group_order <-new_group_order_v1v3_stool$cdf

plot_3 <- plot_microshades(mdf_new_group_order, cdf_new_group_order, group_label = "Phylum Family")

plot_3 + scale_y_continuous(labels = scales::percent, expand = expansion(0)) +
  theme(legend.key.size = unit(0.2, "cm"), text=element_text(size=10)) +
  theme(axis.text.x = element_text(size= 6))

```

### Reorder by both Samples and Abundance of Group levels

```{r fig.width= 8, fig.height= 4}
# reorder_samples_by will change the order of samples based on Family an abundance of "Bacteroidaceae" and reorder the Phylum to sink the most abundant groups
reordered_v1v3_stool <- reorder_samples_by(mdf_v1v3_stool, cdf_v1v3_stool, order = "Bacteroidaceae", group_level = "PHYLUM", subgroup_level = "FAMILY", sink_abundant_groups = TRUE)

mdf_reordered <-reordered_v1v3_stool$mdf
cdf_reordered <-reordered_v1v3_stool$cdf

plot_3 <- plot_microshades(mdf_reordered, cdf_reordered, group_label = "Phylum Family")

plot_3 + scale_y_continuous(labels = scales::percent, expand = expansion(0)) +
  theme(legend.key.size = unit(0.2, "cm"), text=element_text(size=10)) +
  theme(axis.text.x = element_text(size= 6))

```




