---
title: "Human Microbiome Project 2"
author: "Erin Dahl"
date: '`r format(Sys.Date(), "%B %e, %Y")`'
output: rmarkdown::html_document
vignette: >
  %\VignetteIndexEntry{microshades-HMP2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Human Microbiome Project 2 Data Vignette

This vignette uses 16s rRNA sequencing data from the Human Microbiome Project 2.
Vaginal microbiome samples from moms will be examined in this tutorial at the Phylum & Genus taxonomic levels.


This data is available in the HMP2Data library. To download use `BiocManager::install("HMP2Data")`.

Learn more about the HMP2 Data [here](https://bioconductor.org/packages/release/data/experiment/html/HMP2Data.html)


Additionally, the package [speedyseq](https://github.com/mikemc/speedyseq) is necessary to use the function `prep_mdf()`. The package speedyseq provides faster versions of phyloseq’s plotting and taxonomic merging functions. Alternatively, the phyloseq object can be melted and transformed by using phyloseq functions `tax_glom()` and/or `transform_sample_counts()`, and melted by using `psmelt()`.


## Load the necessary packages for this toutorial

```{r message=FALSE, warning=FALSE}
library(HMP2Data)
library(phyloseq)
library(SummarizedExperiment)
library(MultiAssayExperiment)
library(dplyr)
library(ggplot2)
library(microshades)
library(forcats)
library(tidyverse)
```

## Load the HMP2 data as a phyloseq object
```{r}
# Load the momspi16S data 
ps_momspi16S <- momspi16S()

ps_momspi16S
```

The `ps_momspi16S` object contains 9,107 samples. To focus on a smaller sample size, use phyloseq function `subset_samples()`
```{r}
# Subset the samples
ps_momspi16S_sub <- subset_samples(ps_momspi16S,sample_body_site == "vagina")
ps_momspi16S_sub <- subset_samples(ps_momspi16S_sub, visit_number == 9)

ps_momspi16S_sub
```

## Apply microshades functions

Now that the sample size is reduced (22 samples), begin using microshades functions to evaluate abundance and apply advanced color organization. 
```{r}
# Use microshades function prep_mdf to agglomerate, normalize, and melt the phyloseq object
mdf_prep <- prep_mdf(ps_momspi16S_sub)

# Create a color object for the specified data
color_objs_momspi16S <- create_color_dfs(mdf_prep, cvd = TRUE)

# Extract
mdf_momspi16S_unordered <- color_objs_momspi16S$mdf
cdf_momspi16S <- color_objs_momspi16S$cdf
```

### Plot 

The dataframe `mdf_momspi16S_unordered` contains sample data and abundance info. The dataframe `cdf_momspi16S` stores the color mapping information used for plotting. 

`plot_microshades` produces a stacked barplot with ordered subgroup taxonomy. The darkest shade is the most abundant, and the lightest shade is the least abundant. (excluding the "other" subgroup from each of the selected groups)

```{r fig.width= 8, fig.height= 4}
plot_1 <- plot_microshades(mdf_momspi16S_unordered, cdf_momspi16S)

plot_1 + scale_y_continuous(labels = scales::percent, expand = expansion(0)) +
  theme(legend.key.size = unit(0.2, "cm"), text=element_text(size=10)) +
  theme(axis.text.x = element_text(size= 6)) 

```

## Order the samples by Lactobacillus Abundance
The plot above indicates that Lactobacillus is the most abundant genera in the Firmicutes phyla in this dataset. To see samples arranged in order of Lactobacillus abundance (or any genus listed in the legend above)  use the `reorder_samples_by()` function and then plot.

```{r fig.width= 8, fig.height= 4}
# reorder_samples_by will change the order of samples based on an abundance of a specified subgroup taxonomy
# The default subgroup_level is "Genus" and the specified subgroup taxonomy to order by is "Lactobacillus"
momspi16S_ordered <- reorder_samples_by(mdf_momspi16S_unordered,cdf_momspi16S, order = "Lactobacillus")

mdf_momspi16S_ordered <- momspi16S_ordered$mdf
cdf_momspi16S_ordered <- momspi16S_ordered$cdf

plot_2 <- plot_microshades(mdf_momspi16S_ordered, cdf_momspi16S_ordered)

plot_2 + scale_y_continuous(labels = scales::percent, expand = expansion(0)) +
  theme(legend.key.size = unit(0.2, "cm"), text=element_text(size=12)) +
  theme(axis.text.x = element_blank())

```
