---
title: "microshades-GP"
author: "Erin Dahl"
date: '`r format(Sys.Date(), "%B %e, %Y")`'
output: rmarkdown::html_document
vignette: >
  %\VignetteIndexEntry{microshades-GP}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Global Patterns Data Vignette

This vignette uses the Global Patterns microbiome data available from phyloseq.

Learn more about the phyloseq [here](https://github.com/joey711/phyloseq).

Secondary link https://bioconductor.org/packages/release/bioc/html/phyloseq.html

## Load the required packages

```{r warning=FALSE}
library(microshades)
library(phyloseq)
library(ggplot2)
```

## Use the microshades functions

```{r}
# The dataset Global Paterns is a phyloseq object avalible from the Phyloseq package
data(GlobalPatterns)

# Agglomerate and normalize the phyloseq object, and melted to a data frame
mdf_prep <- prep_mdf(GlobalPatterns)

# Generate a color object for the specified data
color_objs_GP <- create_color_dfs(mdf_prep,selected_groups = c("Verrucomicrobia", "Proteobacteria", "Actinobacteria", "Bacteroidetes",
    "Firmicutes") , cvdf = TRUE)

# Extract
mdf_GP <- color_objs_GP$mdf
cdf_GP <- color_objs_GP$color_df
```

### Plot 

Use `mdf_GP` as the object to plot and use `cdf_GP` to assign the correct color assignments. 

```{r fig.width= 9, fig.height= 8}
# Plot
plot <- plot_microshades(mdf_GP, cdf_GP)

plot_1 <- plot + scale_y_continuous(labels = scales::percent, expand = expansion(0)) +
  theme(legend.key.size = unit(0.2, "cm"), text=element_text(size=10)) +
  theme(axis.text.x = element_text(size= 6)) 

plot_1 

```

The `plot_microshades` returns a ggplot object, that allows for additional specifications for the plot to be declared. For example, this allows users to facet samples and more descriptive elements.

```{r eval = FALSE, echo = FALSE}

GP_legend <-custom_legend(cdf_GP, mdf_GP)

GP_legend

```

```{r eval = FALSE, echo = FALSE}

plot_diff <- plot + scale_y_continuous(labels = scales::percent, expand = expansion(0)) +
  theme(legend.position = "none") +
  theme(axis.text.x = element_text(size= 6)) + 
  theme(plot.margin = margin(6,20,6,6))


plot_grid(plot_diff, GP_legend,  rel_widths = c(1, .25))

```

```{r fig.width= 9, fig.height= 4}

plot_2 <- plot + scale_y_continuous(labels = scales::percent, expand = expansion(0)) +
  theme(legend.key.size = unit(0.2, "cm"), text=element_text(size=10)) +
  theme(axis.text.x = element_text(size= 6)) +
  facet_wrap(~SampleType, scales = "free_x", nrow = 2) +
  theme (strip.text.x = element_text(size = 6))

plot_2
```


