---
title: "Sample Reordering"
author: "Anagha Shenoy, Erin Dahl, and Lisa Karstens, PhD"
date: '`r format(Sys.Date(), "%B %e, %Y")`'
output: rmarkdown::html_document
vignette: >
  %\VignetteIndexEntry{sample_reordering}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This tutorial uses the Human Microbiome Project 2 data available from the [`HMP2Data` library](https://bioconductor.org/packages/release/data/experiment/html/HMP2Data.html) to demonstrate the functionality of the `reorder_samples_by` function in microshades.

# Load packages

Load the necessary packages for this tutorial. 

HMP2 data is provided as `phyloseq`, `SummarizedExperiment`, and `MultiAssayExperiment` class objects, so the corresponding packages need to be installed and/or loaded.

```{r message = FALSE, warning=FALSE}

library(microshades)
library(phyloseq)
library(ggplot2)
library(dplyr)
library(cowplot)
library(forcats)
library(tidyverse)

library(HMP2Data)             # BiocManager::install("HMP2Data")
library(SummarizedExperiment) # BiocManager::install("SummarizedExperiment")
library(MultiAssayExperiment) # BiocManager::install("MultiAssayExperiment")

```

# Load and prepare data

```{r}

ps_momspi16S <- momspi16S()

ps_momspi16S

```

The `ps_momspi16S` object contains 9,107 samples. Subset this data to focus on a smaller sample size:

```{r}

ps_momspi16S_sub <- subset_samples(ps_momspi16S, 
                                   sample_body_site == "vagina")

ps_momspi16S_sub <- subset_samples(ps_momspi16S_sub, 
                                   visit_number %in%c(3,9))

```

Here we create a new column with shortened subject IDs, by taking the last three characters of each. 

We will keep samples corresponding to a select list of shortened subject IDs.

These steps are for ease of use in later sections of this tutorial.

```{r}

subj_ids <- sample_data(ps_momspi16S_sub)$subject_id

sample_data(ps_momspi16S_sub)$short_subject_id <- substr(subj_ids, 
                                                         nchar(subj_ids) - 2, 
                                                         nchar(subj_ids))

select_ids <- c("21e", "698", "124", "7f1", 
                "571", "004", "f61", "d96",
                "803", "44e", "a53", "760", 
                "7e1", "e7f", "fc8", "a7d", 
                "dc4", "10e", "fce", "54e", 
                "112")

ps_momspi16S_sub <- subset_samples(ps_momspi16S_sub, short_subject_id %in% select_ids)

```

# Apply microshades functions

Use the `prep_mdf` and `create_color_dfs` microshades functions to evaluate abundance and apply advanced color organization.

```{r}

# Use microshades function prep_mdf to agglomerate, normalize, and melt the phyloseq object
mdf_momspi16S <- prep_mdf(ps_momspi16S_sub,
                          subgroup_level = "Genus")

# Create a color object for the specified data
color_objs_momspi16S <- create_color_dfs(mdf_momspi16S, 
                                         selected_groups = c('Proteobacteria', 'Actinobacteria', 
                                                              'Bacteroidetes', 'Firmicutes'),
                                         group_level = "Phylum",
                                         subgroup_level = "Genus",
                                         cvd = TRUE)

# Extract plotting objects
mdf_momspi16S <- color_objs_momspi16S$mdf
cdf_momspi16S <- color_objs_momspi16S$cdf

```

# Order based on taxon at group level

**Note: in the following examples, `subgroup_level` is Genus and `group_level` is Phylum.**

First, subset to a single visit to focus the view further.

```{r}

ps_momspi16S_v3 <- subset_samples(ps_momspi16S_sub, visit_number == 3)

# Use microshades function prep_mdf to agglomerate, normalize, and melt the phyloseq object
mdf_momspi16S_v3 <- prep_mdf(ps_momspi16S_v3,
                             subgroup_level = "Genus")

# Create a color object for the specified data
color_objs_momspi16S_v3 <- create_color_dfs(mdf_momspi16S_v3,
                                            selected_groups = c('Proteobacteria', 'Actinobacteria', 
                                                                'Bacteroidetes', 'Firmicutes'),
                                            group_level = "Phylum",
                                            subgroup_level = "Genus",
                                            cvd = TRUE)

# Extract plotting objects
mdf_momspi16S_v3 <- color_objs_momspi16S_v3$mdf
cdf_momspi16S_v3 <- color_objs_momspi16S_v3$cdf

```

It is important to note that any specification of the `order_tax` parameter must match either `group_level` or `subgroup_level`, set during creation of the mdf and cdf objects (see above).

Here, we order based on a specific phylum (group level). The following plot is ordered by abundance of the phylum *Actinobacteria*.

```{r}

color_objs_reorder_ab <- reorder_samples_by(mdf_momspi16S_v3,
                                            cdf_momspi16S_v3,
                                            order_tax = "Actinobacteria")

mdf_reorder_ab <- color_objs_reorder_ab$mdf
cdf_reorder_ab <- color_objs_reorder_ab$cdf

```

The `reorder_samples_by` function returns color objects, which must be extracted to then use with the `plot_microshades` function.

```{r fig.width= 6, fig.height= 5}

hmp_legend_1 <- custom_legend(mdf_reorder_ab, cdf_reorder_ab)

hmp_plot_1 <- plot_microshades(mdf_reorder_ab, cdf_reorder_ab) +
  scale_y_continuous(labels = scales::percent, expand = expansion(0)) +
  theme(legend.position = "none") +
  theme(axis.text.x = element_text(size= 8)) +
  theme (strip.text.x = element_text(size = 8))


plot_grid(hmp_plot_1, hmp_legend_1, rel_widths = c(1, .25))

```

# Order based on taxon at subgroup level

Here, we order based on a specific genus (subgroup level). The following plot is ordered by abundance of the specified genus *Lactobacillus*.

```{r}

color_objs_reorder_lb <- reorder_samples_by(mdf_momspi16S_v3,
                                            cdf_momspi16S_v3,
                                            order_tax = "Lactobacillus")

mdf_reorder_lb <- color_objs_reorder_lb$mdf
cdf_reorder_lb <- color_objs_reorder_lb$cdf

```

```{r fig.width= 6, fig.height= 5}

hmp_legend_2 <- custom_legend(mdf_reorder_lb, cdf_reorder_lb)

hmp_plot_2 <- plot_microshades(mdf_reorder_lb, cdf_reorder_lb) +
  scale_y_continuous(labels = scales::percent, expand = expansion(0)) +
  theme(legend.position = "none") +
  theme(axis.text.x = element_text(size= 8)) +
  theme (strip.text.x = element_text(size = 8))


plot_grid(hmp_plot_2, hmp_legend_2, rel_widths = c(1, .25))

```

# Further customization

In addition to reordering samples by abundance of a given taxon, you can supply an ordered list of samples to organize the resulting plot visually.

## Apply alphabetical order

In this example, we would like to sort subject IDs alphabetically. We can achieve this by supplying a list of subject IDs in the desired order to the `sample_ordering` parameter.

```{r}

color_objs_reorder_alpha_subject_ids <- reorder_samples_by(mdf_momspi16S,
                                                           cdf_momspi16S,
                                                           sample_variable = "short_subject_id",
                                                           sample_ordering = sort(select_ids))

mdf_reorder_alpha_ids <- color_objs_reorder_alpha_subject_ids$mdf
cdf_reorder_alpha_ids <- color_objs_reorder_alpha_subject_ids$cdf

hmp_legend_3 <- custom_legend(mdf_reorder_alpha_ids, cdf_reorder_alpha_ids)

hmp_plot_3 <- plot_microshades(mdf_reorder_alpha_ids, cdf_reorder_alpha_ids, x = "short_subject_id") +
  scale_y_continuous(labels = scales::percent, expand = expansion(0)) +
  theme(legend.position = "none") +
  theme(axis.text.x = element_text(size= 8)) +
  facet_wrap(~visit_number, scales = "free_x") +
  theme (strip.text.x = element_text(size = 8)) +
  labs(x = "Subject ID")

plot_grid(hmp_plot_3, hmp_legend_3, rel_widths = c(1, .25))

```

## Order with narrow list

The list supplied to `sample_ordering` does not need to be comprised of *all* samples. 

By default, missing samples in the list will be dropped.

If we only specify the order of samples with shortened subject IDs 004, 10e, 112, 124, and 21e, samples with other subject IDs will be dropped, and a warning will be displayed. (For using microshades in R Markdown files to generate HTMLs, warnings are able to be suppressed by setting `warning=FALSE` in chunk options.)

```{r fig.width=5, fig.height=6}

narrow_list <- c("004", "10e", "112", "124", "21e")

color_objs_reorder_narrow_subject_ids <- reorder_samples_by(mdf_momspi16S,
                                                           cdf_momspi16S,
                                                           sample_variable = "short_subject_id",
                                                           sample_ordering = narrow_list)

mdf_reorder_narrow_ids <- color_objs_reorder_narrow_subject_ids$mdf
cdf_reorder_narrow_ids <- color_objs_reorder_narrow_subject_ids$cdf

hmp_legend_4 <- custom_legend(mdf_reorder_narrow_ids, cdf_reorder_narrow_ids, legend_key_size = 0.8)

hmp_plot_4 <- plot_microshades(mdf_reorder_narrow_ids, cdf_reorder_narrow_ids, x = "short_subject_id") +
  scale_y_continuous(labels = scales::percent, expand = expansion(0)) +
  theme(legend.position = "none") +
  theme(axis.text.x = element_text(size= 8)) +
  facet_grid(vars(visit_number), scales = "free_x") +
  theme (strip.text.x = element_text(size = 8), panel.spacing.y = unit(1.5, "lines")) +
  labs(x = "Subject ID")

plot_grid(hmp_plot_4, hmp_legend_4, rel_widths = c(1, .5))


```


## Apply abundance-based order

The `reorder_samples_by` function can also be used to apply existing abundance-based order to new data. Use `reorder_samples_by` to get the order of a sample variable, then provide this order as a list to reorder another dataset.

In this example, we would like to order by *Lactobacillus* abundance in Visit 3 samples, then apply this order to Visit 9 samples.

First, reorder samples from Visit 3 by *Lactobacillus* abundance, making sure to specify subject ID as the sample variable to sort. The resulting mdf object will be used to extract the order.

```{r}

color_objs_reorder_lb_subject_ids <- reorder_samples_by(mdf_momspi16S_v3,
                                                        cdf_momspi16S_v3,
                                                        sample_variable = "short_subject_id",
                                                        order_tax = "Lactobacillus")

mdf_reorder_lb_ids <- color_objs_reorder_lb_subject_ids$mdf

```

Get the unique factor levels that have been set in the mdf object. Then, convert to a vector to supply to the `reorder_samples_by` function.

``` {r }

lb_desc_ids <- mdf_reorder_lb_ids %>%
                pull(short_subject_id) %>%
                fct_unique() %>%
                as.vector()

```

Reorder by subject ID, following the supplied order of subject IDs. 

```{r }

color_objs_subject_reorder_lb <- reorder_samples_by(mdf_momspi16S,
                                                    cdf_momspi16S,
                                                    sample_variable = "short_subject_id",
                                                    sample_ordering = lb_desc_ids)

mdf_subject_reorder_lb <- color_objs_subject_reorder_lb$mdf
cdf_subject_reorder_lb <- color_objs_subject_reorder_lb$cdf

```

Compare the panels to see how the ordering of *Lactobacillus* abundance in Visit 3 samples has been applied to Visit 9 samples.

```{r }

hmp_legend_6 <- custom_legend(mdf_subject_reorder_lb, cdf_subject_reorder_lb)

hmp_plot_6 <- plot_microshades(mdf_subject_reorder_lb, cdf_subject_reorder_lb, x = "short_subject_id") +
  scale_y_continuous(labels = scales::percent, expand = expansion(0)) +
  theme(legend.position = "none") +
  theme(axis.text.x = element_text(size= 8)) +
  facet_wrap(~visit_number, scales = "free_x", nrow=2) +
  theme (strip.text.x = element_text(size = 8)) +
  labs(x = "Subject ID")

plot_grid(hmp_plot_6, hmp_legend_6, rel_widths = c(1, .25))

```

