---
title: "microshades-GP"
author: "Erin Dahl"
date: '`r format(Sys.Date(), "%B %e, %Y")`'
output: rmarkdown::html_document
vignette: >
  %\VignetteIndexEntry{microshades-GP}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Global Patterns Data Vignette

This vignette uses the Global Patterns microbiome data available from phyloseq.

Learn more about the phyloseq [here](https://github.com/joey711/phyloseq).

Secondary link https://bioconductor.org/packages/release/bioc/html/phyloseq.html

## Load the required packages

```{r warning=FALSE}
library(microshades)
library(phyloseq)
library(ggplot2)
library(dplyr)
library(cowplot)
```

## Use the microshades functions

```{r}
# The dataset Global Patterns is a phyloseq object avalible from the Phyloseq package
data(GlobalPatterns)

# Agglomerate and normalize the phyloseq object, and melted to a data frame
mdf_prep <- prep_mdf(GlobalPatterns)

# Generate a color object for the specified data
color_objs_GP <- create_color_dfs(mdf_prep,selected_groups = c("Verrucomicrobia", "Proteobacteria", "Actinobacteria", "Bacteroidetes",
    "Firmicutes") , cvdf = TRUE)

# Extract
mdf_GP <- color_objs_GP$mdf
cdf_GP <- color_objs_GP$color_df
```

### Plot 

Use `mdf_GP` as the object to plot and use `cdf_GP` to assign the correct color assignments. 

```{r fig.width= 9, fig.height= 8}
# Plot
plot <- plot_microshades(mdf_GP, cdf_GP)

plot_1 <- plot + scale_y_continuous(labels = scales::percent, expand = expansion(0)) +
  theme(legend.key.size = unit(0.2, "cm"), text=element_text(size=10)) +
  theme(axis.text.x = element_text(size= 6)) 

plot_1 

```

The `plot_microshades` returns a ggplot object, that allows for additional specifications for the plot to be declared. For example, this allows users to facet samples and more descriptive elements.


```{r fig.width= 9, fig.height= 4}

plot_2 <- plot + scale_y_continuous(labels = scales::percent, expand = expansion(0)) +
  theme(legend.key.size = unit(0.2, "cm"), text=element_text(size=10)) +
  theme(axis.text.x = element_text(size= 6)) +
  facet_wrap(~SampleType, scales = "free_x", nrow = 2) +
  theme (strip.text.x = element_text(size = 6))

plot_2
```

Use custom legend Format
```{r fig.width= 9, fig.height= 4}

GP_legend <-custom_legend(mdf_GP, cdf_GP)

plot_diff <- plot + scale_y_continuous(labels = scales::percent, expand = expansion(0)) +
  theme(legend.position = "none")  +
  theme(axis.text.x = element_text(size= 6)) +
  facet_wrap(~SampleType, scales = "free_x", nrow = 2) +
  theme(axis.text.x = element_text(size= 6)) + 
  theme(plot.margin = margin(6,20,6,6))

plot_grid(plot_diff, GP_legend,  rel_widths = c(1, .25))

```

Re-examine data with smaller groups
```{r}
# Subset Global Patterns to smaller groups in the dataset
ps_water <- subset_samples(GlobalPatterns, SampleType %in% c("Freshwater", "Freshwater (creek)", "Ocean"))

ps_land <- subset_samples(GlobalPatterns, SampleType %in% c("Soil", "Sediment (estuary)"))

ps_human <- subset_samples(GlobalPatterns, SampleType %in% c("Skin", "Feces", "Tongue"))

# Agglomerate and normalize the phyloseq objects, and melt to a data frame
mdf_water <- prep_mdf(ps_water)
mdf_land <- prep_mdf(ps_land)
mdf_human <- prep_mdf(ps_human)

# Generate a color object for the specified data
color_objs_water <- create_color_dfs(mdf_water,selected_groups = c("Verrucomicrobia", "Proteobacteria", "Actinobacteria", "Bacteroidetes",
    "Firmicutes") , cvdf = TRUE)

color_objs_water <- reorder_samples_by(color_objs_water$mdf, color_objs_water$color_df)

color_objs_land <- create_color_dfs(mdf_land,selected_groups = c("Verrucomicrobia", "Proteobacteria", "Actinobacteria", "Bacteroidetes",
    "Firmicutes") , cvdf = TRUE)

color_objs_land <- reorder_samples_by(color_objs_land$mdf, color_objs_land$color_df)

color_objs_human <- create_color_dfs(mdf_human,selected_groups = c("Verrucomicrobia", "Proteobacteria", "Actinobacteria", "Bacteroidetes",
    "Firmicutes") , cvdf = TRUE)

color_objs_human <- reorder_samples_by(color_objs_human$mdf, color_objs_human$color_df)

# Extract
mdf_water <- color_objs_water$mdf
cdf_water <- color_objs_water$color_df

mdf_land <- color_objs_land$mdf
cdf_land <- color_objs_land$color_df

mdf_human <- color_objs_human$mdf
cdf_human <- color_objs_human$color_df
```

Water Samples 
```{r}
water_legend <-custom_legend(mdf_water, cdf_water, legend_key_size = .3, legend_text_size = 8)

water_plot <- plot_microshades(mdf_water, cdf_water) + 
  scale_y_continuous(labels = scales::percent, expand = expansion(0)) +
  theme(legend.position = "none") +
  theme(axis.text.x = element_text(size= 6)) +
  facet_wrap(~SampleType, scales = "free_x") +
  theme (strip.text.x = element_text(size = 6)) 


plot_grid(water_plot, water_legend,  rel_widths = c(1, .25))


```

Land samples

```{r}

land_legend <-custom_legend(mdf_land, cdf_land, legend_key_size = .3, legend_text_size = 8)

land_plot <- plot_microshades(mdf_land, cdf_land) + 
  scale_y_continuous(labels = scales::percent, expand = expansion(0)) +
  theme(legend.position = "none") +
  theme(axis.text.x = element_text(size= 6)) +
  facet_wrap(~SampleType, scales = "free_x") +
  theme (strip.text.x = element_text(size = 6)) 


plot_grid(land_plot, land_legend,  rel_widths = c(1, .25))

```


Human Samples
```{r}

human_legend <-custom_legend(mdf_human, cdf_human, legend_key_size = .3, legend_text_size = 8)

human_plot <- plot_microshades(mdf_human, cdf_human) + 
  scale_y_continuous(labels = scales::percent, expand = expansion(0)) +
  theme(legend.position = "none") +
  theme(axis.text.x = element_text(size= 6)) +
  facet_wrap(~SampleType, scales = "free_x") +
  theme (strip.text.x = element_text(size = 6)) 


plot_grid(human_plot, human_legend,  rel_widths = c(1, .25))

```
